---
title: "Probability Practice Assignment"
author: "Haden Loveridge, Alex Parson, Biagio Alessandrello"
date: "2024-08-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Probability Practice

__Part A.__ 

1. Define the probabilities and given data:
- P(RC) = 0.3 #Random Clickers
- P(TC) = 1 - P(RC) = 0.7 #Truthful Clickers
- P(Yes|RC) = 0.5
- (Yes) = 0.65
2. Rule of total probability (What fraction of people who are truthful clickers answered yes)
$$ P(A) = \sum_{i=1}^n P(A|B_i)P(B_i)$$ 
$$P(Yes) = P(Yes|RC)P(RC) + P(Yes|TC)P(TC)$$
- 0.65 = (0.5 * 0.3) + P(Yes|TC) * 0.7
- 0.65 = 0.15 + P(Yes|TC) * 0.7
- 0.50 = P(Yes|TC) * 0.7
- P(Yes|TC) = 0.7143

Fraction of people who are truthful clickers that answered yes is 71.43%

__Part B.__ 

1. Define the probabilities and given data:
- P(Positive Test|Has Disease) = 0.993
- P(Negative Test|Doesn't Have Disease).= 0.9999
- P(Has Disease) = 0.000025
- P(Doesn't Have Disease) = 1 - P(Has Disease) = 0.999975
2. What is the probability that hey have the disease if the person tests positive?
$$P(Has Disease|Positive Test)$$
$$P(\text{Has Disease}|\text{Positive Test}) = \frac{P(\text{Positive Test}|\text{Has Disease}) \cdot P(\text{Has Disease})}{P(\text{Positive Test})}$$ 
- First find P(Positive Test) using rule of total probabilities
$$ P(\text{Positive Test}) = P(\text{Positive Test}|\text{Has Disease}) \cdot P(\text{Has Disease}) + P(\text{Positive Test}|\text{Doesnt Have Disease}) \cdot P(\text{Doesnt Have Disease}) $$
-   P(Positive Test) = 0.0001248
- Use Bayes Theorem
-   P(Has Disease|Positive Test) = 0.1988

The probability that a person who tested positive actually has the disease is around 19.88%

```{r Math, include=FALSE}
p_positive_test_has_disease = 0.993
p_negative_test_no_disease = 0.9999
p_has_disease = 0.000025
p_no_disease = 1 - p_has_disease
p_positive_test = p_positive_test_has_disease * p_has_disease + (1 - p_negative_test_no_disease) * p_no_disease
p_has_disease_positive_test = (p_positive_test_has_disease * p_has_disease) / p_positive_test
```







